A Hierarchical Bayesian Framework for Stellar Obliquity Distribution Inference

### 2023-02-11 src/tex/ms.tex
[showyourwork] overleaf sync

Stellar obliquity $\psi$, the angle between a planet's orbital axis and its host star's spin axis, traces the formation and evolution of a planetary system. In transiting-exoplanet observations, only the sky-projected stellar obliquity $\lambda$ can be measured. To find the true stellar obliquity $\psi$, information about stellar inclination $i_\star$ is also required.
Here we show that, while the stellar inclination is important to infer the stellar obliquity of an individual system, they are not necessarily required for the population stellar obliquity inference. 
Constraints on the stellar obliquity distribution are predominantly contributed by the sky-projected stellar obliquities.
We introduce a flexible, hierarchical Bayesian framework that could infer the stellar obliquity distribution purely from sky-projected stellar obliquities. Stellar inclination measurements are optional in the model, and if not provided, are assumed to be isotropically distributed.
Applying the framework to all exoplanetary systems with sky-projected stellar obliquity measurements, we find that the stellar obliquity distribution is unimodal and peaked at zero degrees. Misaligned systems have nearly isotropic stellar obliquities, which may have important implications for the formation and evolution of planetary systems.

@@

It is yet unclear if all of these physical and dynamic processes apply to exoplanetary systems. These proposed mechanisms have different predictions on stellar obliquity distributions \citep[see][and references therein]{Albrecht22}. For example, in the secular chaos mechanism, the stellar obliquity could never get beyond $90\degr$ \citep{Teyssandier19}. In the stellar Kozai mechanism, the stellar obliquity distribution should be bimodal and concentrated at $40\degr$ and $140\degr$ \citep[e.g.,][]{Fabrycky07, Anderson16, Vick19}. In the multiple-planet scattering mechanism, the majority of systems should be aligned, and a small fraction of systems have a wide range of stellar obliquities \citep[e.g.,][]{Beague12}. Motivated by these predictions, we aim to constrain the stellar obliquity distribution of exoplanetary systems from a Bayesian approach to find dominant mechanisms that sculpt close-in planetary systems.

In exoplanetary observation, only the sky-projected stellar obliquity $\lambda$, the angle between the projections of $\bf{n}_{\rm orb}$ and $\bf{n}_{\star}$ onto the plane of the sky, can be measured. Stellar obliquity $\psi$ can be calculated, if the stellar inclination $i_\star$ is known. The relation follows \citep[e.g.,][]{Fabrycky09}:
\begin{equation}\label{eqn:psi}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ is the angle between the vector $\bf{n}_{\rm orb}$ and the observer's line of sight, and $i_\star$ is the angle between $\bf{n}_{\star}$ and the observer's line of sight.
For transiting exoplanet systems where $i_{\rm orb} \approx 90\degr$, $\cos{\psi} \approx \sin{i_\star}\cos{\lambda}$.

Stellar inclination can often be constrained via, e.g., photometric and spectroscopic rotational modulation introduced by starspots, gravity darkening, and asteroseismology \citep[see][and references therein]{Albrecht22}.
For systems without $i_\star$ measurements, it is still possible to infer their stellar obliquities from the sky-projected obliquities, assuming isotropic stellar inclinations. The inferred $\psi$ will have greater uncertainty than the one inferred with $i_\star$ measurement \citep{Fabrycky09}.

We do not yet understand the importance of stellar inclination in population stellar obliquity inference.
In this work, we infer the stellar obliquity distribution from sky-projected stellar obliquities and show that the stellar obliquity distribution weakly depends on stellar inclinations.
In Section~\ref{sec:hbm}, we introduce a flexible, hierarchical Bayesian framework to infer the population stellar obliquity distribution.
In Section~\ref{sec:applications}, we apply the framework to simulated data and real observations.
In Section~\ref{sec:jacobian}, we find the mathematical relation between the $\psi$, $\lambda$, and $i_\star$ distributions and explain why stellar inclination is not important in stellar obliquity distribution inference.

>>>>

Stellar obliquity $\psi$, the angle between a planet's orbital axis and its host star's spin axis, traces the formation and evolution of a planetary system. In transiting-exoplanet observations, only the sky-projected stellar obliquity $\lambda$ can be measured. To find the true stellar obliquity $\psi$, information about stellar inclination $i_\star$ is also required.
Here we show that, while the stellar inclination is important to infer the stellar obliquity of an individual system, they are not necessarily required for the population stellar obliquity inference. 
Constraints on the stellar obliquity distribution are predominantly contributed by the sky-projected stellar obliquities.
We introduce a flexible, hierarchical Bayesian framework that could infer the stellar obliquity distribution purely from sky-projected stellar obliquities. Stellar inclination measurements are optional in the model, and if not provided, are assumed to be isotropically distributed.
Applying the framework to all exoplanetary systems with sky-projected stellar obliquity measurements, we find that the stellar obliquity distribution is unimodal and peaked at zero degrees. Misaligned systems have nearly isotropic stellar obliquities, which may have important implications for close-in planets' formation and evolution.

@@

It is yet unclear if all of these physical and dynamic processes apply to exoplanetary systems. These proposed mechanisms have different predictions on stellar obliquity distributions \citep[see][and references therein]{Albrecht22}. For example, in the secular chaos mechanism, the stellar obliquity could never get beyond $90\degr$ \citep{Teyssandier19}. In the stellar Kozai mechanism, the stellar obliquity distribution should be bimodal and concentrated at $40\degr$ and $140\degr$ \citep[e.g.,][]{Fabrycky07, Anderson16, Vick19}. In the multiple-planet scattering mechanism, the majority of systems should be aligned, and a small fraction of systems have a wide range of stellar obliquities \citep[e.g.,][]{Beague12}. Motivated by these predictions, we characterize the stellar obliquity distribution of exoplanetary systems from a Bayesian approach to find the dominant mechanisms that sculpt close-in planetary systems.

In exoplanetary observation, only the sky-projected stellar obliquity $\lambda$, the angle between the projections of $\bf{n}_{\rm orb}$ and $\bf{n}_{\star}$ onto the plane of the sky, can be measured, mainly via the Rossiter-McLaughlin effect \citep{Rossiter24, McLaughlin24}. The stellar obliquity $\psi$ can be calculated, if both the sky-projected stellar obliquity $\lambda$ and the stellar inclination $i_\star$ are known. The relation follows \citep[e.g.,][]{Fabrycky09}:
\begin{equation}\label{eqn:psi}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ is the inclination angle between the vector $\bf{n}_{\rm orb}$ and the observer's line of sight, and $i_\star$ is the inclination angle between $\bf{n}_{\star}$ and the observer's line of sight.
For transiting exoplanet systems where $i_{\rm orb} \approx 90\degr$, $\cos{\psi} \approx \sin{i_\star}\cos{\lambda}$.

Stellar inclinations can often be constrained via, e.g., photometric and spectroscopic rotational modulation introduced by starspots \citep[e.g.,][]{Masuda20, Albrecht21}, gravity darkening \citep[e.g.,][]{Barnes09, Barnes11}, and asteroseismology \citep[e.g.][]{Chaplin13}.
For systems without $i_\star$ measurements, it is still possible to infer their stellar obliquities from the sky-projected obliquities, assuming isotropic stellar inclinations; however, the inferred $\psi$ will have greater uncertainty than the one inferred with $i_\star$ measurement \citep{Fabrycky09}.

We do not yet understand the role of stellar inclination in population stellar obliquity inference.
In this work, we study the relationship between the distributions of stellar obliquity, sky-projected stellar obliquity, and stellar inclination, and introduce a statistical approach to infer the stellar obliquity distribution.
In Section~\ref{sec:hbm}, we introduce a flexible, hierarchical Bayesian framework that infers the stellar obliquity distribution of a sample.
In Section~\ref{sec:applications}, we apply the framework to simulated data and real observations and show that the stellar obliquity distribution inferred purely from sky-projected stellar obliquities will not compromise our interpretation.
In Section~\ref{sec:jacobian}, we find the mathematical relation between the $\psi$, $\lambda$, and $i_\star$ distributions and explain why the stellar obliquity distribution inference is predominantly determined by sky-projected stellar obliquities.
